Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
updates

\clearpage
\hrule
I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
\begin{itemize}
\item simple linear regression $y=\beta_0 + \beta_1x_1$
\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
\end{itemize}
So RF, DL and LR are all multi-variable (in this application).

A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
\hrule

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we apply both the supervised LR and unsupervised \kMeans\ method on the full dataset of 14 populations and assess the 
classification accuracy into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
%DENIS: above you argue that kMeans has local optima and results might hence be too optimistic but here you only train one model. So either remove the above section or add some repeats to this one
This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).

@@

Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
RF is building an forest of decision trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.

LR is

\marginpar{TODO [write up the observations and argue why one is less resource hungry than the other one]}

\begin{table}
\caption{Runtime and memory consumption on the 4 super-populations}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & nodes & runtime & memory \\
RF  & X (Y) & X (Y) & X (Y) \\
\kMeans & X (Y) & X (Y) & X (Y) \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the 
%unsupervised \kMeans{} approach or the 
multivariate random forest or deep learning approach as reported by
Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
of the 1000 genome project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately identify all individuals from EUR (lowest precision) or misclassify AMR (lowest recall) individuals. 
We therefore ague that the remaining mixture is potentially reflective of the migrational background, which may be an important observation for medical
applications that need accurate population association.

>>>>

%\clearpage
%\hrule
%I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
%\begin{itemize}
%\item simple linear regression $y=\beta_0 + \beta_1x_1$
%\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
%\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
%\end{itemize}
%So RF, DL and LR are all multi-variable (in this application).

%A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
%\hrule

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
The ARI allows us to compare the classification results with \kMeans{} clustering results, as it takes the partitioning of the dataset into account, rather than just labels.
report the mean error over all training and testing folds
along with the standard error of the mean (SEM)
We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we apply both the supervised LR and unsupervised \kMeans\ method on the full dataset of 14 populations and assess the 
classification accuracy into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).

@@

Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
RF is building an forest of decision trees.
When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
(6GB for RF). Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, at the expense of time. 
However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability.
The memory requirements also restricts a job of that size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
be an essential step for future work, especially as datasets grow outwards.

\begin{table}
\caption{Runtime and memory consumption on the 4 super-populations}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & chromosomes & executors & runtime & memory (GB) \\
\kMeans & 22 & 80 & 2mins 3secs & 3 \\
LR  & 22 & 80 & 5mins 1sec & 3 \\
RF & 22 & 80 & 17mins 20secs & 6 \\
\kMeans & 1-22 & 40 & 14hrs 44mins & 24 \\
LR  & 1-22 & 40 & 30hrs 12mins & 24 \\
\noalign{\smallskip}
\hline
\label{resources}
\end{tabular}
\end{center}
\end{table}

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the 
%unsupervised \kMeans{} approach or the 
multivariate random forest or deep learning approach as reported by
Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1096 individuals
of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately identify all individuals from EUR (lowest precision) or misclassify AMR (lowest recall) individuals. 
We therefore ague that the remaining mixture is potentially reflective of the migrational background, which may be an important observation for medical
applications that need accurate population association.
